---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(dplyr)
```
# gendercoder
<!-- badges: start -->

[![CRAN status](https://www.r-pkg.org/badges/version/gendercoder)](https://CRAN.R-project.org/package=gendercoder)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/ropensci/gendercoder/workflows/R-CMD-check/badge.svg)](https://github.com/ropensci/gendercoder/actions)
[![Codecov test coverage](https://codecov.io/gh/ropensci/gendercoder/branch/master/graph/badge.svg)](https://codecov.io/gh/ropensci/gendercoder?branch=master)[![ROpenSci
status](https://badges.ropensci.org/435_status.svg)](https://github.com/ropensci/software-review/issues/435)
<!-- badges: end -->


The goal of gendercoder is to allow simple re-coding of free-text gender 
responses. This is intended to permit representation of gender diversity, 
while managing troll-responses and the workload implications of manual coding. 

## Installation

This package is not on CRAN. To use this package please run the following code:

``` r
devtools::install_github("ropensci/gendercoder")
library(gendercoder)
```
## Basic use

The gendercoder package permits the efficient re-coding of free-text gender 
responses within a tidyverse pipeline. It contains two built-in English output 
dictionaries, a default `manylevels_en` dictionary which corrects spelling and 
standardises terms while maintaining the diversity of responses and a 
`fewlevels_en` dictionary which contains fewer gender categories, "man", "woman", 
"boy", "girl", and "sex and gender diverse". 

The core function, `gender_recode()`, takes 3 arguments, 

- `gender` the vector of free-text gender,

- `dictionary` the preferred dictionary, and

- `retain_unmatched` a logical indicating whether original values should be carried over if 
there is no match. 

Basic usage is demonstrated below. 

```{r}
library(gendercoder)

tibble(gender = c("male", "MALE", "mle", "I am male", "femail", "female", "enby")) %>% 
  mutate(manylevels_gender  = recode_gender(gender, dictionary = manylevels_en, retain_unmatched = TRUE),
         fewlevels_gender = recode_gender(gender, dictionary = fewlevels_en, retain_unmatched = FALSE)
  )
  
```

The package does not need to be used as part of a tidyverse pipeline:

```{r}
df <- tibble(gender = c("male", "MALE", "mle", "I am male", "femail", "female", "enby")) 

df$manylevels_gender <- recode_gender(df$gender, dictionary = manylevels_en)
df
```


## Contributing to this package

This package is a reflection of cultural context of the package contributors. 
We acknowledge that understandings of gender are bound by both culture and time 
and are continually changing. As such, we welcome issues and pull requests to 
make the package more inclusive, more reflective of current understandings of 
gender inclusive languages and/or suitable for a broader range of cultural 
contexts. We particularly welcome addition of non-English dictionaries or of 
other gender-diverse responses to the `manylevels_en` and `fewlevels_en` dictionaries.

The ["Adding to the dictionary"](https://ropensci.github.io/gendercoder/articles/a02_add_dictionary.html) vignette includes information about how to make changes to the dictionary either for your own use or when contributiong to the gendercoder package.

## Citation Information

Please cite this package as:

Jennifer Beaudry, Emily Kothe, Felix Singleton Thorn, Rhydwyn McGuire, Nicholas Tierney and Mathew Ling (2020).
  gendercoder: Recodes Sex/Gender Descriptions into a Standard Set. R package version 0.0.0.9000.
  https://github.com/ropensci/gendercoder

## Acknowledgement of Country

We acknowledge the Wurundjeri people of the Kulin Nation as the custodians of 
the land on which this package was developed and pay respects to elders past, 
present and future.
