---
title: Using a custom dictionary"
author: "Emily Kothe"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This gendercodeR package allows rapid recoding of freetext gender responses using an in-built dictionary. Individuals using the package may wish to extend or overwrite elements of the dictionary where they disagree with the coding decisions made by the package authors and/or wish to include recoding options that are not included in the in-built dictionary. A custom dictionary can be created by the user and provided to the `genderRecode()` function for this purpose. 

## Adding new items to the dictionary

For example, when collecting data from Spanish speaking participants a participant who enters their gender as "masculino" will not be recoded using the in-built dictionary.

```{r}

df <- data.frame(stringsAsFactors=FALSE,
      gender = c("male", "MALE", "mle", "mail", "masculino", "female", "enby"),
         age = c(34L, 37L, 77L, 52L, 68L, 67L, 83L)
)

df

library(gendercodeR)

genderRecoded <- genderRecode(df)
```

The custom dictionary should be a data frame with two columns. Column 1 should include the input text (i.e. the response provided by the participant) and Column 2 should include the desired output text (i.e. the gender category the input should be converted to).

```{r}

custom_dictionary <- data.frame(stringsAsFactors=FALSE,
       Input = c("masculino", "femenina"),
      Output = c("male", "female")
)

custom_dictionary
```

The custom dictionary is appended to the in-built dictionary and participants are recoded as desired.

```{r}
genderRecode(df, customDictionary = custom_dictionary)
```

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
