---
title: "rationale"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rationale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# gendercodeR

The goal of gendercodeR is to allow simple re-coding of free-text gender responses.

## Installation

This package is not on CRAN. To use this package please run the following code:

``` r
devtools::install_github("ropenscilabs/gendercodeR")
library(gendercodeR)
```

## Why would we do this?

Researchers who collect self-reported demographic data from respondents 
occasionally collect gender using a free-text response option. This has the 
advantage of respecting the gender diversity of respondents without prompting 
users and potentially including misleading responses. However, this presents a 
challenge to researchers in that some inconsistencies in typography and spelling 
create a larger set of responses than would be required to fully capture the 
demographic characteristics of the sample. 

For example, male participants may provide free-text responses as "male", "man", 
"mail", "mael". Non-binary participants may provide responses as "nonbinary", 
"enby", "non-binary", "non binary"

Manually coding of such free-text responses this is often not feasible with 
larger datasets. `gendercodeR()` uses dictionaries of common 
misspellings to re-code free-text responses into a consistent set of 
responses. The small number of responses not automatically re-coded by 
gendercodeR() can then be feasibly manually recoded. 

## Motivating example

`gendercodeR()` includes a sample dataset with actual free-text 
responses to the question "What is your gender?" from a number of studies of 
English-speaking participants. The sample dataset includes responses from 7756
participants. Naive coding identifies 103 unique responses to this item. 


```{r message = FALSE, warning= FALSE}
library(gendercodeR)
library(dplyr)

sample %>% 
  group_by(Gender) %>% 
  summarise(count = n()) %>% 
  arrange(-count) %>% 
  knitr::kable(caption = "Summary of gender categories before coding")
```

Recoding using the `gender_coder()` function classifies all but 28 responses 
into pre-defined response categories. 

```{r}
sample %>% 
  head(10) %>% 
  mutate(recoded_gender = recode_gender(gender = Gender, dictionary = broad)) %>% 
  knitr::kable(caption = "The broad dictionary applied to `head(sample)`")

sample %>% 
  mutate(recoded_gender = recode_gender(gender = Gender, dictionary = broad)) %>% 
  filter(!is.na(recoded_gender)) %>% 
  group_by(recoded_gender) %>% 
  summarise(count = n()) %>% 
  arrange(-count) %>% 
  knitr::kable(caption = "Summary of gender categories after use of the *broad* dictionary")

```

In this dataset unclassified responses are a mix of unusual responses and 
apparent response errors (e.g. numbers and symbols). While some of these are
genuinely missing (i.e. Gender = 40), other could be manually recoded, or added
to a custom dictionary. 

```{r}

sample %>% 
  mutate(recoded_gender = recode_gender(gender = Gender, dictionary = broad)) %>% 
  filter(is.na(recoded_gender)) %>% 
  knitr::kable(caption = "All responses not classified by the inbuilt dictionary")

```

## Options within the function

### dictionary

The package provides two inbuilt dictionaries. The use of these is controlled 
using the `dictionary` argument. The first `dictionary = broad` provides 
corrects spelling and standardises terms while maintaining the diversity of responses. 
This is the default dictionary for gendercodeR() as it preserves as much gender
diversity as possible. 

However in some cases you may wish to collapse gender into a smaller set of 
categories by using the `narrow` dictionary (`dictionary = narrow`). This 
dictionary contains only three gender categories, "male", "female", and 
"sex and gender diverse". 

The "male" category includes all participants who indicate that they are

- male
- trans male  (including female to male transgender respondents)
- cis male

The "female" category includes all participants who indicate that they are

- female
- trans female (including male to female transgender respondents)
- cis female

The "sex and gender diverse" category includes all participants who indicate 
that they are

- agender
- androgynous
- intersex
- non-binary
- gender-queer

```{r} 
sample %>% 
  head(10) %>% 
  mutate(recoded_gender = recode_gender(gender = Gender, dictionary = narrow)) %>% 
  knitr::kable(caption = "The narrow dictionary applied to `head(sample)`")

sample %>% 
  mutate(recoded_gender = recode_gender(gender = Gender, dictionary = narrow)) %>% 
  group_by(recoded_gender) %>% 
  summarise(count = n()) %>% 
  arrange(-count) %>% 
  knitr::kable(caption = "Summary of gender categories after use of the *narrow* dictionary")
```

You can also specify a custom dictionary to replace or supplement the inbuilt 
dictionary. The custom dictionary should be a list in the following 
format. 

```{r}
# name of the list element is the user input value and the list element is the 
# replacement value corresponding to that name as a lower case string.
custom_dictionary <- list(
  masculino = "male",
  hombre = "male",
  mujer = "female",
  femenina = "female"
)

str(custom_dictionary)
```

Custom dictionaries can be used in place of an inbuilt dictionary or can 
supplement the inbuilt dictionary by providing a vector of lists to the 
dictionary argument. Where the lists contain duplicated elements the last 
version of the duplicated value will be used for re-coding. This allows you to 
use the inbuilt dictionary but change the coding of one or more responses from 
that dictionary. Here the addition of Spanish terms allows for recoding of 11 
previously uncoded responses.

```{r}

sample %>% 
  mutate(recoded_gender = recode_gender(gender = Gender, 
                                        dictionary = c(narrow, 
                                                       custom_dictionary))) %>% 
  group_by(recoded_gender) %>% 
  summarise(count = n()) %>% 
  arrange(-count) %>% 
  knitr::kable(caption = "Summary of gender categories after use of the combined dictionaries")


sample %>% 
  mutate(recoded_gender = recode_gender(gender = Gender, 
                                        dictionary = c(narrow, 
                                                       custom_dictionary))) %>% 
  filter(is.na(recoded_gender)) %>% 
  knitr::kable(caption = "All responses not classified by the combined dictionaries")
```

### fill

The `fill` argument is used to determine the handling the re-coding of values
not contained in the dictionary. By default, unmatched values are coded as NA. 
`fill = TRUE` will fill unmatched responses with the participant provided 
response. 

```{r}
sample %>% 
  mutate(recoded_gender = recode_gender(gender = Gender, 
                                        dictionary = c(narrow, 
                                                       custom_dictionary),
                                        fill = TRUE)) %>% 
  group_by(recoded_gender) %>% 
  summarise(count = n()) %>% 
  arrange(-count) %>% 
  knitr::kable(caption = "Summary of gender categories after use of the combined dictionary and `fill = TRUE`")
```

# A disclaimer on gender

This package attempts to remove typographical errors from free text gender data.
The defaults that we used are specific to our context and your data may be
different. 

We offer two builtin dictionaries, board and narrow. Both are opinionated
about how gender descriptors collapse into categories. 

However, as these are culturally specific they may not be suitable for your data. 
In particularly the narrow setting makes opinionated choices about some responses 
that we want to acknowledge are potentially problematic. In particular,

- In 'narrow' coding intersex responses are recoded as 'sex and gender
diverse'
- In 'narrow' responses where people indicate they are trans and
indicate their identified gender are recoded as the identified gender
(e.g. 'Male to Female' is recoded as Female). We wish to acknowledge
that this may not reflect how some individuals would classify
themselves when given these categories and in some contexts may make
systematic errors. The broad coding dictionary attempts to avoid these
issues as much as possible - however users can provide a custom
dictionary to add to or overwrite our coding decisions if they feel
this is more appropriate. We welcome people to update the inbuilt
dictionary where desired responses are missing.
- In both dictionaries we assume that typographical features such as spacing
are not relevant to re-coding of the gender response (e.g. we assume that
"genderqueer" and "gender queer" are equivalent). This is unlikely to be 
true for all contexts. 


The 'broad' coding separates out those who identify as trans
female/male or cis female/male into separate categories it should not
be assumed that all people who describe as male/female are cis, if you
are assessing trans status we recommend a two part question see:

Bauer, Greta & Braimoh, Jessica & Scheim, Ayden & Dharma, Christoffer.
(2017). Transgender-inclusive measures of sex/gender for population surveys:
Mixed-methods evaluation and recommendations. PLoS ONE. 12.

# Contributing to this package

This package is a reflection of cultural context of the package contributors we 
welcome issues and pull requests to make the package more inclusive and/or 
suitable for a broader range of cultural contexts. We particularly welcome 
addition of non-English dictionaries or of other genderdiverse responses to the 
broad and narrow dictionaries.

# Acknowledgement of Country

We acknowledge the Wurundjeri people of the Kulin Nation as the custodians of 
the land on which this package was developed and pay respects to elders past, 
present and future.
